NumPy---Array Computation for Python

### 2020-01-04 body.tex
Use two columns
Reorganize text

With its first stable release in October 2006, NumPy unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
NumPy consists of the \code{ndarray}-object along with utility functions that
operate on it.
Because of its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for tensor computation---witnessed by its use in popular
deep learning libraries such as PyTorch\cite{pytorch}.

\subsection{Background}

@@

The first incarnation of Numpy was a package called "Numeric", written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Huginin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.

As Hugunin points out, at the time he wrote Numeric, commercial languages for
interactive computing with numerical arrays were already well-established,
including Matlab and IDL.  At the time, it must have seemed a safe bet that
Numeric would not succeed in this market.  The commercial languages were
developed and maintained by large numbers of full-time professional programmers
and supported by formal quality assurance, release managers and writers of
scientific documentation.  In contrast, Numeric, and later, Numpy, would
continue to be developed almost entirely by scientists, including graduate
students and faculty, without specific funding, and often to the detriment of
the scientists' careers.

@@

Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily, and, around 2000, wrote a re-implementation of much of
Numeric, called Numarray, to support their need to work on large memory-mapped
arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and Numarray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.  At the time, Oliphant was an assistant
professor of Electrical and Computer Engineering at Brigham Young University.

The creation of NumPy spurred renewed development in the larger scientific
Python ecosystem and heralded the current era of wide-spread use of Python for
scientific computing.

Over the next 10 years, Numpy attracted a relatively large pool of scientist
developers (see Figure ~\ref{fig:prs-over-time}).  To benefit from these
user-developers, the project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improved collaboration on code, and
continuous testing so we run a large battery automated tests for every proposed
change in Numpy. At first, Numpy had rather sparse documentation, but the project
developed a new standard for ``docstrings'' —- text describing each NumPy
function -- and used this standard to write comprehensive high-quality
documention, integrated with the source code. The Numpy documentation format
quickly became a standard for the larger Python community.

As we implied above, there were several factors that allowed this rapid growth
and successful development.

@@

Python is already a good choice for many standard programming tasks such as
cleaning data, interacting with web resources and parsing text.  It has a wide
range of libraries for many different tasks. Adding fast array operations and
linear algebra allows the scientist to do all their work with within the same
language - and one that has the advantage of being famously easy to learn and
teach.

The second factor flows from Python's nature as an open-source language,
embedded within the open-source community.  Python attracts developers who like
to build and contribute.  As a result, Numpy has always been a library that was
developed by scientists in order to do their work.  Commercial specialist
numerical languages can encourage consumer-users, who pick up a tool and use
it, but do not expect to contribute to development.  In contrast, Numpy often
attracts user-developers, who want to work with Numpy, and are prepared to fix
and extend the library as they work. User-developers have been the fundamental
force in developing Numpy, and shaping it into a library that has been refined
in the furnace of practical scientific work.

@@

Lastly, we believe Numpy has benefitted greatly from a well-chosen application
programming interface (API).  Scientific and numerical programming needs to be
fast, and scale to very large datasets.  The Numpy API defines a simple wrapper
for data in memory so it can be represented as a one- or multi-dimensional
array.  The simplicity of this wrapper has made it very successful as a
standard way of representing arrays in memory, and has made it relatively easy
for other libraries to develop fast and memory-efficient compiled code, usually
in C or Fortran, that can manipulate these arrays and pass them back to Python.
This has been a large factor in the growth of numerical computing libaries
around Numpy and Python.

@@

\subsection{Funding}

@@

\section{Community and library}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
    \caption{Number of pull requests merged into the NumPy master branch for each
        quarter since 2012. The total number of PRs is indicated with the
        lower blue area showing the portion contributed by current or previous
        maintainers.}\label{fig:prs-over-time}
\end{figure}

\subsection{Basic functionality}

@@

\fixme{discuss Figure~\ref{fig:basic-functionality} .. vectors / matrices / and then more ...
indexing and slicing, one-liners that would be tens or hundreds of lines of code in
languages such as C.}

@@

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
\begin{lstlisting}
>>> import numpy as np
>>> x = np.array([0, 1, 2, 3])
>>> x.dtype
dtype('int64')
>>> x.shape
(4,)

@@

>>> x @ y
28
\end{lstlisting}
  \caption{Vectors (1-d arrays)}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
\begin{lstlisting}
>>> import numpy as np
>>> x = np.array([1, 2])
>>> A = np.array([[0.1, 0.2], [0.3, 0.2]])
>>> A.dtype 
dtype('float64')
>>> A.shape
(2, 2)

@@

>>> A @ x
array([0.5, 0.7])
>>> x @ A
array([0.7, 0.6])
\end{lstlisting}
  \caption{Vectors (1-d arrays) and Matrices (2-d arrays)}
  \label{fig:sub2}
\end{subfigure}
\caption{Basic functionality}
\label{fig:basic-functionality}
\end{figure}

@@

\subsection{Advanced functionality}

The \emph{universal functions} (or \emph{ufuncs})
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-d array with length 3, and
\code{y} is a 2-d array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$ (see Figure~\ref{fig:broadcasting}).  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to compute the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.

\begin{figure}
  \centering
  \includegraphics[width=0.25\textwidth]{static/broadcasting}
  \caption{
    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
    yields a $2 \times 3$ result.
  }
  \label{fig:broadcasting}
\end{figure}

NumPy also provides basic infrastructure for other packages in the scientific
Python ecosystem.
The \code{numpy.testing} subpackage provides functions such as
\code{assert\_allclose(actual, desired)} that may be used in unit
test suites for code that uses NumPy arrays.
The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
BLAS/LAPACK, and other relevant libraries for scientific computing.
The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

@@

The NumPy codebase has been actively developed for over two decades (first as
Numeric and NumArray).
Given the complexity of the codebase and the massive number of projects depending
on it, large changes require careful planning and substantial work.
As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently and informally until 2017 when,
to facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
%Since then there have been 19 proposed NEPS---6 have been implemented,
%4 have been accepted and are being implemented, 4 are under
%consideration, 3 have been deferred or superseded, and 2 have been rejected
%or withdrawn.

@@

x_gpu = cp.array([1, 2, 3])
y = np.sum(x_gpu)  # This works, and returns a GPU array
\end{lstlisting}

Similarly, user defined functions composed using NumPy can now be
applied to, e.g., multi-node distributed Dask arrays:

\begin{lstlisting}
import numpy as np
import dask.array as da

def f(x):
    """Function using NumPy API calls"""
    y = np.tensordot(x, x.T)
    return np.mean(np.log(y + 1))

x_local = np.random.random([10000, 10000])  # random local array
x_distr = da.random.random([10000, 10000])  # random distributed array

f(x_local)  # returns a NumPy array
f(x_distr)  # works, returns a Dask array
\end{lstlisting}

@@

\begin{lstlisting}
from numpy.random import Generator, PCG64, SeedSequence

seq = SeedSequence(10304245474441179933310169597522947680)
pcg = PCG64(seq)
gen = Generator(pcg)
\end{lstlisting}

>>>>

@@

\subsection{History}

The first incarnation of NumPy was a package called "Numeric", written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Hugunin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.

As Hugunin points out, at the time he wrote Numeric, commercial languages for
interactive computing with numerical arrays were already well-established,
including Matlab and IDL.  At the time, it must have seemed a safe bet that
Numeric would not succeed in this market.  The commercial languages were
developed and maintained by large numbers of full-time professional programmers
and supported by formal quality assurance, release managers and writers of
scientific documentation.  In contrast, Numeric, and later, NumPy, would
continue to be developed almost entirely by scientists, including graduate
students and faculty, without specific funding, and often to the detriment of
the scientists' careers.

@@

Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily, and, around 2000, wrote a re-implementation of much of
Numeric, called NumArray, to support their need to work on large memory-mapped
arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.  At the time, Oliphant was an assistant
professor of Electrical and Computer Engineering at Brigham Young University.

Over the next 10 years, NumPy attracted a relatively large pool of scientist
developers.  To benefit from these
user-developers, the project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improved collaboration on code, and
continuous testing so we run a large battery automated tests for every proposed
change in NumPy. At first, NumPy had rather sparse documentation, but the project
developed a new standard for ``docstrings''---text describing each NumPy
function---and used this standard to write comprehensive high-quality
documentation, integrated with the source code. The NumPy documentation format
quickly became a standard for the larger Python community.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
    \caption{Number of pull requests merged into the NumPy master branch for each
        quarter since 2012. The total number of PRs is indicated with the
        lower blue area showing the portion contributed by current or previous
        maintainers.}\label{fig:prs-over-time}
\end{figure}

\subsection{Enabling factors}

The creation of NumPy spurred renewed development in the larger scientific
Python ecosystem and heralded the current era of wide-spread use of Python for
scientific computing.
As we implied above, there were several factors that allowed this rapid growth
and successful development.

@@

Python is already a good choice for many standard programming tasks such as
cleaning data, interacting with web resources and parsing text.  It has a wide
range of libraries for many different tasks. Adding fast array operations and
linear algebra allows the scientist to do all their work with within the same
language---and one that has the advantage of being famously easy to learn and
teach.

The second factor flows from Python's nature as an open-source language,
embedded within the open-source community.  Python attracts developers who like
to build and contribute.  As a result, NumPy has always been a library that was
developed by scientists in order to do their work.  Commercial specialist
numerical languages can encourage consumer-users, who pick up a tool and use
it, but do not expect to contribute to development.  In contrast, NumPy often
attracts user-developers, who want to work with NumPy, and are prepared to fix
and extend the library as they work. User-developers have been the fundamental
force in developing NumPy, and shaping it into a library that has been refined
in the furnace of practical scientific work.

@@

Lastly, we believe NumPy has benefited greatly from a well-chosen application
programming interface (API).  Scientific and numerical programming needs to be
fast, and scale to very large datasets.  The NumPy API defines a simple wrapper
for data in memory so it can be represented as a one- or multi-dimensional
array.  The simplicity of this wrapper has made it very successful as a
standard way of representing arrays in memory, and has made it relatively easy
for other libraries to develop fast and memory-efficient compiled code, usually
in C or Fortran, that can manipulate these arrays and pass them back to Python.
This has been a large factor in the growth of numerical computing libraries
around NumPy and Python.

@@

\subsection{Hybrid development}

@@

\section{The \code{ndarray} data structure}

With its first stable release in October 2006, NumPy unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
NumPy consists of the \code{ndarray}-object along with utility functions that
operate on it.
Because of its inherent simplicity---being a pointer to memory with some
associated meta-data about shape, data-type, and so forth---the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for tensor computation---witnessed by its use in popular
deep learning libraries such as PyTorch\cite{pytorch}.

@@

\fixme{discuss .. vectors / matrices / and then more ...
indexing and slicing, one-liners that would be tens or hundreds of lines of code in
languages such as C.}

@@

\begin{lstlisting}
>>> import numpy as np
>>> x = np.array([0, 1, 2, 3])
>>> x.dtype
dtype('int64')
>>> x.shape
(4,)

@@

>>> x @ y
28
\end{lstlisting}

\begin{lstlisting}
>>> import numpy as np
>>> x = np.array([1, 2])
>>> A = np.array([[0.1, 0.2], [0.3, 0.2]])
>>> A.dtype 
dtype('float64')
>>> A.shape
(2, 2)

@@

>>> A @ x
array([0.5, 0.7])
>>> x @ A
array([0.7, 0.6])
\end{lstlisting}

@@

%\subsection{Advanced functionality}
%
%The \emph{universal functions} (or \emph{ufuncs})
%are functions written in C that implement efficient looping over
%NumPy arrays. An important feature of ufuncs is the built-in
%implementation of \emph{broadcasting}.  For example, the function
%\code{arctan2(x, y)} is a ufunc that accepts two arguments and computes
%$\tan^{-1}(y/x)$.  The ufunc loops over the dimensions of the inputs
%in such a way that if, say, \code{x} is a 1-d array with length 3, and
%\code{y} is a 2-d array with shape $2 \times 1$, the output will be
%an array with shape $2 \times 3$ (see Figure~\ref{fig:broadcasting}).  The ufunc machinery takes care
%of calling the function with all the appropriate combinations of
%input array elements to compute the output array.
%The elementary arithmetic operations of addition, multiplication, etc.,
%are implemented as ufuncs, so broadcasting applies even in expressions
%such as \code{x + y * z}.
%
%\begin{figure}
%  \centering
%  \includegraphics[width=0.5\linewidth]{static/broadcasting}
%  \caption{
%    Broadcasting a $1 \times 3$ array against a $2 \times 1$ array
%    yields a $2 \times 3$ result.
%  }
%  \label{fig:broadcasting}
%\end{figure}
%
%NumPy also provides basic infrastructure for other packages in the scientific
%Python ecosystem.
%The \code{numpy.testing} subpackage provides functions such as
%\code{assert\_allclose(actual, desired)} that may be used in unit
%test suites for code that uses NumPy arrays.
%The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
%BLAS/LAPACK, and other relevant libraries for scientific computing.
%The program \code{f2py} is a tool for
%building NumPy-aware Python wrappers of Fortran functions.
%NumPy itself does not use any Fortran code;  F2PY is part of NumPy
%for historical reasons.

@@

Given the complexity of the codebase and the massive number of projects depending
on it, large changes require careful planning and substantial work.
As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently and informally until 2017 when,
to facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
%Since then there have been 19 proposed NEPS---6 have been implemented,
%4 have been accepted and are being implemented, 4 are under
%consideration, 3 have been deferred or superseded, and 2 have been rejected
%or withdrawn.

@@

x_gpu = cp.array([1, 2, 3])
y = np.sum(x_gpu)  # Returns a GPU array
\end{lstlisting}

%
%Similarly, user defined functions composed using NumPy can now be
%applied to, e.g., multi-node distributed Dask arrays:
%
%\begin{lstlisting}
%import numpy as np
%import dask.array as da
%
%
%def f(x):
%    """Function using NumPy API calls"""
%    y = np.tensordot(x, x.T)
%    return np.mean(np.log(y + 1))
%
%
%x_local = np.random.random([10000, 10000])  # random local array
%x_distr = da.random.random([10000, 10000])  # random distributed array
%
%f(x_local)  # returns a NumPy array
%f(x_distr)  # works, returns a Dask array
%\end{lstlisting}

@@

\begin{lstlisting}
from numpy.random import (
    Generator,
    PCG64,
    SeedSequence,
)

seq = SeedSequence(1030424547444117993331016959)
pcg = PCG64(seq)
gen = Generator(pcg)
\end{lstlisting}
